Positivity-Preserving Adaptive Runge--Kutta Methods

### 2021-03-01 nusslein_vsrp/latex/response_to_reviewers.tex

\begin{document}
\begin{letter}{}

\textbf{Response to Reviewer}

\opening{Dear Editor, Dear Reviewer,}

We are very grateful for the thorough review and helpful comments on our manuscript
``Positivity-Preserving Adaptive Runge--Kutta Methods''.
We revised the manuscript in accordance with the reviewer's suggestions.
Please find below a point-by-point description of these changes and our reply to
the reviewer's comments.

@@

\begin{enumerate}[label=\arabic*.]
  \setcounter{enumi}{1}
  \item Bound-preserving adaptive Runge–Kutta methods
  \begin{enumerate}[label=\alph*)]
    \item \emph{%
    In equation (2.14b), it is weird to read $-0.01 u_2$ and $- 0.05 u_2$
    in the same line. Is it an error? The same thing happens in equations (2.14c)
    with $u_3$, in (5.5b) with $u_2$, and in (5.5c) with $u_3$.
    }

    We added an explanation of we use this unusual form of the
    equations [lines~\ref{line:prod-dest}ff.].

@@

  \item Selection of the modified weights
  \begin{enumerate}[label=\alph*)]
    \item \emph{%
    In section 3.2, two different approaches to adapt the weights are considered:
    Free adaptation and Convex adaptation. At the same time, it is explained that
    ``...we can additionally impose either or both of the following strategies \dots''.
    I think that the first strategy is nothing but the second approach (Convex
    adaptation). I suggest making this part clearer in order to avoid confusion
    between approaches and strategies.
    }

    We modified this part to be more clear [lines~\ref{line:strategies}ff.].
  \end{enumerate}

  \item Properties of adaptive RKMs
  \begin{enumerate}[label=\alph*)]
    \item \emph{%
    Figure 3a, with free adaptation of the weights $\tilde b$, does not give enough
    information about the stability regions of the adapted methods. In my opinion,
    it would be clearer if each stability curve was labeled with the value of
    the difference $\| \tilde b - b \|$, accordingly with the values in Figure 4b.
    In this way we could see the progress of the stability regions during the
    numerical integration process.
    }

    To indicate the value of $\| \tilde b - b \|$ for the different stability
    regions, we now color the boundaries using a colormap and provide a colorbar.
	We also removed the filling pattern to improve clarity.

    Additionally, we discovered that the step size used for this plot was not
    the correct value ($\Delta t = 0.015$) corresponding to Figure 4a. We
    corrected the step size. Because of this,
    the shape of the plotted stability regions changed in the revision.

@@

    \item \emph{%
    After formula (5.2) ``...In Figure 4a the results for $\Delta t = 0.015$
    are plotted for different time steps...'' should be ``...In Figure 4a the
    results for $\Delta t = 0.015$ are plotted for different values of time $t$...''
    }

    We fixed this sentence [line~\ref{line:time-vs-dt}].

    \item \emph{%
    Figure 5, about the non-stiff problem with fixed stepsize, is quite interesting
    and gives information about the convergence of the adapted mehtod. For
    $\Delta t = 0.0082$ the unaltered method leads to positive solutions. For
    $\Delta t = 0.015$ the original method leads to negative values in the
    interval $[0, 0.375]$ and the weights are altered by solving an LP problem.
    Is this extra time consuming a significant cost or could it be an economical
    alternative to using excessively small step sizes? A further explanation
    about this issue should be added.
    }

    We investigated this and added a remark at the end of Section 5.1, explaining
    that the LP approach is more expensive in this case but would hopefully be
    more efficient for certain more challenging problems.  [line~\ref{line:cost}]

### 2021-03-01 nusslein_vsrp/latex/overview.tex
Rephrase some of the revisions.

\subsection{Non-stiff problem with fixed stepsize}\label{sec:Ex_expl}
First, adaptive RKMs based on explicit methods are tested on non-stiff problems with a fixed step size.
When used with explicit methods the cost of solving the LP is significant because the computation of the stage derivatives only requires $s$ evaluations of the
\linelabel{line:RHS}\revA{right-hand side (RHS)}.
For most of the linear test problems tried, the explicit methods yield to positive results.
When increasing the step size, issues with stability occur before getting negative values.
An example for this is the ODE in Section~\ref{sec:example_lin}.
Some nonlinear RHS may require very small time steps to preserve positivity.
For these, adapting the weights could be a possible way to solve them.
An example is the reaction equation solved in Section~\ref{sec:example_reac}.
\linelabel{line:RHS-negative-values}\revA{%
Since the positivity of the stage values is not ensured, the RHS has to be defined
for negative values. If this is not the case, it to be adapted accordingly.
For example, an RHS involving a square root can throw an error (forcing the
numerical algorithm to terminate) or return ``not a number'' (\texttt{NaN})
when called with negative values, depending on the implementation.
In both cases, further (useful) computations are not possible. A possible fix
allowing our mechanism to continue the integration can be to restrict the arguments
to the square root to non-negative values by clipping them accordingly, e.g.\
to replace \texttt{sqrt(u)} by \texttt{sqrt(max(u, 0))} or something similar.}

>>>>

\subsection{Non-stiff problem with fixed stepsize}\label{sec:Ex_expl}
First, adaptive RKMs based on explicit methods are tested on non-stiff problems with a fixed step size.
When used with explicit methods the cost of solving the LP is significant because the computation of the stage derivatives only requires $s$ evaluations of the
\linelabel{line:RHS}\revA{right-hand side (RHS)}.
For most of the linear test problems tried, the explicit methods yield to positive results.
When increasing the step size, issues with stability occur before getting negative values.
An example for this is the ODE in Section~\ref{sec:example_lin}.
Some nonlinear RHS may require very small time steps to preserve positivity.
For these, adapting the weights could be a possible way to solve them.
An example is the reaction equation solved in Section~\ref{sec:example_reac}.
\linelabel{line:RHS-negative-values}\revA{%
Since the stage values are not guaranteed to be positive, the RHS must be
defined also
for negative values. If there is not a natural definition for negative values,
one can instead extend the function in a smooth way or simply replace negative
stage values by zero,  e.g.\ replace \texttt{sqrt(u)} by \texttt{sqrt(max(u, 0))}.}
